---
title: 'brolgar: An R package to BRowse Over Longitudinal Data Graphically and Analytically in R'
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
  pdf_document:
    keep_tex: yes
---

<!-- --- -->
<!-- title: "brolgar: An R package to BRowse Over Longitudinal Data Graphically and Analytically in R" -->
<!-- author: -->
<!--   # see ?rjournal_article for more information -->
<!--   - name: Nicholas Tierney -->
<!--     affiliation: Monash University -->
<!--     address: -->
<!--     - Monash Clayton -->
<!--     url: https://njtierney.com -->
<!--     orcid: 0000-0002-9079-593X -->
<!--     email:  nicholas.tierney@gmail.com -->
<!--   - name: Dianne Cook -->
<!--     url: https://dicook.org -->
<!--     email: dicook@monash.edu -->
<!--     orcid: 0000-0002-9079-593X -->
<!--     affiliation: Monash University -->
<!--     address: -->
<!--     - Monash Clayton -->
<!--     affiliation2: ACEMS -->
<!--     address2: -->
<!--     - ACEMS Brisbane -->
<!-- abstract: > -->
<!--   An abstract of less than 150 words. -->
<!-- preamble: | -->
<!--   % Any extra LaTeX you need in the preamble -->

<!-- # per R journal requirement, the bib filename should be the same as the output  -->
<!-- # tex file. Don't forget to rename the bib file and change this example value. -->
<!-- bibliography: brolgar-paper.bib -->
<!-- output: rticles::rjournal_article -->
<!-- --- -->


```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r libraries, message = FALSE}
library(tidyverse)
library(kableExtra)
library(lme4)
library(modelr)
library(tsibble)
library(brolgar)
library(patchwork)
library(gridExtra)
library(conflicted)
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")
conflict_prefer("pull", "dplyr")
conflict_prefer("heights", "brolgar")
heights <- brolgar::heights %>% filter(year > 1700)
# NOTE - how to discuss that the data has been filtered from original source?
```

# Abstract

Longitudinal (panel) data provide the opportunity to examine temporal patterns of individuals, because measurements are collected on the same person at different, and often irregular, time points. The data is typically visualised using a "spaghetti plot", where a line plot is drawn for each individual. When overlaid in one plot, it can have the appearance of a bowl of spaghetti. With even a small number of subjects, these plots are too overloaded to be read easily. The interesting aspects of individual differences are lost in the noise. Longitudinal data is often modeled with a hierarchical linear model to capture the overall trends, and variation among individuals, while accounting for various levels of dependence. However, these models can be difficult to fit, and can miss unusual individual patterns. Better visual tools can help to diagnose longitudinal models, and better capture the individual experiences. This paper introduces the R package, brolgar (BRowse over Longitudinal data Graphically and Analytically in R), which provides tools to identify and summarise interesting individual patterns in longitudinal data.

# Introduction

This paper is about exploring longitudinal data effectively. Longitudinal data provide the opportunity to examine temporal patterns of individuals, because measurements are collected on the same individual at different (often irregular), time points. Put another way, longitudinal data can be defined as individuals repeatedly measured through time. Figure \@ref(fig:heights-sample-plot) show a table and plot of sample data of the average height of Australian males. The individual component here `country`, and the time component is `year`. The variable `country` is measured repeatedly from 1900 to 1970, with irregular time periods between years.


```{r heights-sample-plot, fig.cap = "A sample of average heights of men in Australia for 1900-1970. The heights increase over time, but are measured at irregular intervals."}

heights_oz <- heights %>% 
  filter(country == "Australia") %>% 
  slice_tail(n = 5) 

mytheme <- gridExtra::ttheme_default(
  core = list(padding = unit(c(5, 10), "mm"))
  )

tbl_heights_oz <- heights_oz %>% 
  mutate(height_cm = round(height_cm, 1)) %>% 
  tableGrob(theme = mytheme, 
            rows = NULL)

gg_heights_oz <- ggplot(heights_oz,
       aes(x = year,
           y = height_cm,
           group = country)) + 
  geom_point() + 
  geom_line()
# NOTE - add time points for each measurement?

grid.arrange(tbl_heights_oz,
             gg_heights_oz,
             nrow = 1,
             as.table = TRUE)
```

The full data of Figure \@ref(fig:heights-sample-plot) is shown in Figure \@ref(fig:spaghetti), of 144 countries. The data is shown from the year 1700. This plot is challenging to understand because there is overplotting, which makes it hard to see the individuals. Showing separate individual plots of each country doesn't help to understand the data; 144 plots is too many. Making the lines transparent or fitting a simple model to all the data Figure \@ref(fig:spaghetti)B, might be a common first step to see common trends. However, all this seems to clarify is: 1) some set of countries that are more similar, whose masss lies in the center, and 2) a general upward trend in heights over time. We learn about the group, but lose sight of the individuals. 

Some common tasks in exploration are not straightforward: How to look at samples of the data? How to find interesting samples? How to find interesting individuals, finding who has the greatest change in height? Which has the tallest or shortest people? Which are most similar to the min, max, and median?

```{r spaghetti, fig.height = 4, fig.width = 12, message = FALSE}
spaghetti <- ggplot(heights,
       aes(x = year,
           y = height_cm,
           group = country)) + 
  geom_line()  +
  labs(tag = "A")

spaghetti_alpha <- ggplot(heights,
       aes(x = year,
           y = height_cm,
           group = country)) + 
  geom_line(alpha = 0.3) +
  labs(tag = "B")

spaghetti_smooth <- spaghetti_alpha + geom_smooth(method = "lm", 
                                                  aes(group = NULL)) +
  labs(tag = "C")

spaghetti + spaghetti_alpha + spaghetti_smooth
```

```{r spaghetti-interesting-obs}
# code to identify: 
# For example, that the country with the greatest change in heights is XX, XX is the country with the tallest people, and XX has shortest people (Figure XX).
```

This paper demonstrates efficient exploration of longitudinal data, by examining four problems in exploring longitudinal data: 

1. How to look at some sample of the data
2. How to find interesting individuals
3. Finding representative individuals
4. How to understand a model

This paper proceeds in the following way: first, a brief review of the definition of longitudinal data, and existing approaches to longitudinal data, then approaches to these four problems are discussed, followed by a summary.

# Longitudinal Data Structures

There are many names of longitudinal data: Panel data, survey data, repeated measures, and time series, to name a few. Although there might be small differences among these definitions, referring to data collection methods, context, and field, these data structures all share a fundamental similarity. They are measurements over some time period.

To efficiently work with longitudinal data, we need to define it's structure in such a way that cuts across these different definitions. Doing so provides a consistent way to understand and summarise the data. As defined earlier, longitudinal data can be thought of as "individuals repeatedly measured through time". This means that we can consider longitudinal data a type of time series, which is defined in @fpp: "Anything that is observed sequentially over time **is a time series**".

This definition has been realised in data as a time series `tsibble` in [@Wang2018]. These objects are defined as data that have these conditions met:

1. The `index`: the time variable
2. The `key`: variable(s) defining individual groups (or series)
3. 1 + 2 together determine a distinct row

If the specified key and index pair do not define a distinct row - for example, if there are duplicates in the data, the `tsibble` will not be created. This helps ensure the data is properly understood and cleaned up before analysis is conducted, removing avoidable errors that might have impacted downstream decisions.

We can formally define our `heights` data from Figure \@ref(fig:heights-sample-plot) as a `tsibble`: The `index` is `year`, the `key` is `country`. Defining a `tsibble` object is perfomed using `as_tsibble` below, specifying `regular = FALSE`, since there are not regular intervals in the years measured:


```{r define-heights, eval = FALSE}
heights <- as_tsibble(heights,
                      index = year,
                      key = country,
                      regular = FALSE)

```

Using a `tsibble` means this important time series information is recorded **once**, and can be referred to many times in other parts of the data analysis. 

The `tsibble` package also works with `dplyr` family, extending them to provide functions to explore implicit missing values in the `index` (e.g., `has_gaps()` and `fill_gaps()`), as well as grouping and partitioning based on the index with `index_by()`. For more details see @Wang2018.

The benefits to using `tsibble` are that it works within the `tidyverse` ecosystem, in addition there are a tidy series of packages for time series called "tidyver**ts**", containing `fable` [@fable], `feasts`, [@feasts]. The `brolgar` package uses `tsibble` so users can take advantage of these tools and have as few extra pieces as possible.

# Other approaches to longitudinal data

The `pmdplyr` package provides "Panel Manouvers" in "dplyr". It defines the data structure in roughly analagous terms to the `index` and `key` in `tsibble`, but instead requires `id` and `group`. It mostly concerns doing efficient and custom joins, using terminology from econometrics, such as ... it provides its own data structure that is effectively equivalent to `tsibble`, but does not implement 

`panelr` provides an interface focussed on model fitting in regression, such as ...
It also provides widening and lengthening functions. This package fits a different niche to `brolgar`, instead providing helpers on modelling, and it's own interfaces into models. This is in opposition to brolgar, which provides instead workflow focussed helpers.

R provides basic time series, `ts`, objects, which are vectors or matrices that represent data sampled at equally spaced points in time. These have been extended through a variety of other packages, such as `xts`, and `zoo`, which only consider data in a wide format with a regular implied time series. These are not appropriate for longitudinal data, which can have indexes that are not time unit oriented, such as "Wave 1...2", etc, or may contain irregular intervals.

`plm`

Most of these packages provide more than just data definitions, they all share similar features, that there is some ID / index and some group / key feature. 

These software generally reimplement their own custom panel data class object, as well as custom data cleaning tasks, such as reshaping into long and wide form. However, operations for custom long and wide form are not needed since they are handled within `tsibble` + `dplyr`. There are indeed some useful aspects of the specific data cleaning, but they are limited in how they interact with other R packages. `brolgar` provides more flexibility since it relies on `tsibble`, which ensures data manipulation with `dplyr` respects groups. Additionally, it is not limited to specific panel data tools, and can work more generally with general modelling software, in particular other ecosystems such as `fable`  and `feasts`. 

A lot of this is predicated on the fact that longitudinal data, panel data, repeated measures, and all those variations on the names of this dataset, can be thought of as being a type of time series. This is a unifying feature that helps bring the wider toolset from time series into the world longitudinal + panel data world.

https://cran.r-project.org/web/packages/pmdplyr
https://nickch-k.github.io/pmdplyr/index.html
https://github.com/jacob-long/panelr
https://panelr.jacob-long.com/articles/wbm.html
https://bookdown.org/ccolonescu/RPoE4/panel-data-models.html
http://www.mysmu.edu/faculty/anthonytay/Notes/Panel_in_R.pdf
https://cran.r-project.org/web/packages/plm/index.html
https://cran.r-project.org/web/packages/plm/vignettes/plmPackage.html


# How to look at some sample of the data

Plots like Figure \@ref(fig:spaghetti) are often called, "spaghetti plots", and can be useful for a high level understand of the dataset. However, we cannot process and understand the individuals. Just how spaghetti is portioned out for consumption, we can portion out these plots using a set of small subplots to create a plot like Figure \@ref(fig:facet-sample). 

```{r facet-sample, fig.cap = "Facet Samples"}
ggplot(heights,
       aes(x = year,
           y = height_cm,
           group = country)) + 
  geom_line() + 
  facet_sample()
```


`brolgar` provides the `facet_sample()` function, which means the user only needs to consider the most relevant questions: "How many keys per facet?" and "How many facets to look at?", defaulting to 12 facets and three keys per facet. The code to create this figure is shown with:

```r
ggplot(heights,
       aes(x = year,
           y = height_cm,
           group = country)) + 
  geom_line() + 
  facet_sample()
```

Extending this idea of samples, we can instead look at all of the data spread out equally over facets, using `facet_strata()`:

```{r facet-strata, echo = TRUE}
ggplot(heights,
       aes(x = year,
           y = height_cm,
           group = country)) + 
  geom_line() + 
  facet_strata()
```

Here each key is shown only once, and is randomly assigned to a facet/subplot. We can provide meaning to each of these facets, by arranging the heights along a variable, like `year`:

```{r facet-year-along, echo = TRUE}
heights %>% 
ggplot(aes(x = year,
           y = height_cm,
           group = country)) + 
  geom_line() + 
  facet_strata(along = -year)
```

Not that we have not lost any of the data, only the order they are presented in has changed. We learn the distribution and changes in heights over time, and that those measured from the earliest times appear to be more similar, but there is much wider variation in the middle years, and then for more recent heights measured from the early 1900s, the heights are more similar. The starting point of each of these years seems to increase at roughly the same interval. This tells us that the starting times of the years is roughly uniform.

Together `facet_sample()` and `facet_strata()` allow for rapid exploration, by focussing on relevant questions instead of minutiae on details of appropriately randomly assigning while maintaining key structure, keeping the correct number of keys per plot, and so on. For example, `facet_sample()` the questions are: "How many lines per facet" and "How many facets?", and for `facet_strata()` the questions are: "How many facets / strata?" and "What to arrange plots along?". 

Answering these questions keeps the analysis in line with the analytic goals of exploring the data, rather than distracting away by moving attention to minutiae. This is a key theme of improving tools for data analysis. Abstracting away the parts that are not needed, so the analyst can focus on the task at hand.

Under the hood, `facet_sample()` and `facet_strata()` are powered with `sample_n_keys()` and `stratify_keys()`, so the user can use these to create the data structures needed for these plots, and extend them to other purposes that we haven't considered.


We have addressed "how to look at some of the data" by using `tsibble` to store important data on the key and index components, which in turn allow us to look at many and all subsamples using `facet_sample()` and `facet_strata()`.

# Exploring measurement interval

Longitudinal data is not always measured at the same time and at the same frequency.  

We can explore when countries are first getting measured using `features` function from the `fabletools` package, which takes a `tsibble` object and calculates a summary for each `key`. It provides a summary of a given variable, accounting for the time series structure, and returning one row per key specified. It can be thought of as a time-series aware variant of `dplyr::summarise`. We can find the first year for each country number of starting years like so:

```{r feature-first}
heights %>% 
  features(year, list(first = first))
```

And then plot the first year like so: 

```{r feature-first-gg}
heights %>% 
  features(year, list(first = first)) %>% 
  ggplot(aes(x = first)) +
  geom_bar() # data is already binned into 10 year blocks.
```

We can explore the variation in first year using `feat_diff_summary`. This combines many summaries of the differences in `year`.

```{r heights-diff-summary}
heights_diffs <- heights %>% 
  features(year, feat_diff_summary)

heights_diffs
```

This is particularly useful, since `diff` will return a very wide dataset:

```{r diff-wide}
heights %>% 
  features(year, diff)
```

We can then look at the summaries of the differences in year by changing to long form and then facetting:

```{r heights-long-feat-diff}
heights_diff_long <- heights_diffs %>% 
  pivot_longer(cols = -country,
               names_to = "feature",
               values_to = "value") %>% 
  mutate(feature = factor(feature,
                          levels = c("diff_min",
                                     "diff_q25",
                                     "diff_median",
                                     "diff_mean",
                                     "diff_q75",
                                     "diff_max",
                                     "diff_sd",
                                     "diff_var")))

ggplot(heights_diff_long,
       aes(x = value)) +
  geom_histogram(binwidth = 10,
                 colour = "white") +
  facet_wrap(~feature,
             scales = "free")
```

We learn that the smallest interval between measurements is 10 years, and the largest interval is between 10 and 125 years, and that most of the data is measured between 10 and 30 or so years.

# How to find interesting individuals

A workflow to identify interesting individuals starts with: "what defines interesting?". In this case, let us suggest interesting is "tallest height", or "maximum height".

The workflow is represented in the figure XX below, which shows:

1. Identify features based on "interesting" metric (in this case max height)
2. This results in one observation per key
3. Identify from these features what is most interesting (e.g., the highest max height)
4. Filter the most interesting
5. Join this feature back to the data to get all observations for those keys
6. These can now be shown with the data, or separately.

(this will be a figure that combines slides 56 - 69 of https://njt-rstudio20.netlify.app/#85)

These features of interest are identified using `features`. We can calculate our own "feature", calculating the maximum height like so:

```{r heights-max}
heights %>% 
  features(height_cm, list(min = min))
```

Where we pass `features` the variable we want to calculate a feature on, and a named list of features. You can use a prepared summary, for which `brolgar` provides a few - we've already seen `feat_diff_summary`, which provided a summary of the `diff` of a variable. `feat_ranges` produces range related features, min, max, difference from the range, and the interquartile range.

```{r heights-features}
heights %>%
  features(height_cm, feat_ranges)

```

`brolgar` provides some useful additional features for the five number summary, `feat_five_num`, whether keys are monotonically increasing `feat_monotonic`, and measures of spread or variation, `feat_spread`. What is particularly great is that these will work on any type of time series data, and you can use other more classical time series features from `feasts`, such as autocorrelation, `feat_acf()`, Seasonal and Trend decomposition using Loess `feat_stl()` [@ref].

This demonstrates a workflow that can be used to understand and explore your longitudinal data. `brolgar` builds upon this workflow made available by `feasts` and `fabletools`. Users can also create their own features to summarise the data.

# How to understand a model

We now demonstrate how these workflows can be used to interpret and explore a model. Let's take an example of the fixed effect of year + random intercept for country like so:

```{r heights-lmer}
heights_fit <- lmer(height_cm ~ year + (1|country), brolgar::heights)
heights_aug <- brolgar::heights %>%
  add_predictions(heights_fit, var = "pred") %>%
  add_residuals(heights_fit, var = "res")
```

We can use the previous approach to explore the model results. First, looking at all the data (Figure \@ref(fig:gg-heights-lmer)A), then looking at many subsamples with `facet_sample()` (Figure \@ref(fig:gg-heights-lmer)B), all subsamples with `facet_strata()` (Figure \@ref(fig:gg-heights-lmer)C), all subsamples, arranging along residuals, with `facet_strata(along = -res)` \@ref(fig:gg-heights-lmer)D). 

(Not: I'm not entirely sure how useful this is? I mean the last one is kinda useful?)

```{r gg-heights-lmer, fig.cap = "Figure of"}
gg_heights_aug <- ggplot(heights_aug,
       aes(x = year,
           y = pred,
           group = country)) + 
  geom_line() 

p1 <- gg_heights_aug + labs(tag = "A")

p2 <- gg_heights_aug + facet_sample(n_facets = 6) + labs(tag = "B")

p3 <- gg_heights_aug + facet_strata(n_strata = 6) + labs(tag = "C")

p4 <- gg_heights_aug + facet_strata(n_strata = 6, along = -res) + labs(tag = "D")

p1 / p2 / (p3 + p4)
```

We can take a look at a sample of the predictions along with the data, by using `sample_n_keys`. This provides a useful way to explore some set of the model predictions.
In order to find those predictions that best summarise the best, and worst, and in between, we need to use the methods in the next section, "finding representative individuals".

```{r sample-n-keys-model, fig.cap = "Exploration of a random sample of the data against the predictions"}
heights_aug %>% 
  sample_n_keys(10) %>% 
  ggplot(aes(x = year,
             y = pred,
             group = country)) + 
  geom_line() +
  geom_point(aes(y = height_cm)) + 
  facet_wrap(~country)
```

# Finding representative individuals

```{r heights-aug-res-summary, fig.cap = "Five number summary of residual values"}
heights_aug_res_summary <- 
tibble(names = names(summary(heights_aug$res)),
       values = as.numeric(round(summary(heights_aug$res),2))) %>% 
  slice(-4) %>% 
  mutate(y_pos = c(280, 275, 275, 250, 275))

gg_resid <- 
ggplot(heights_aug,
       aes(x = res)) + 
  geom_histogram(colour = "white",
                 binwidth = 1)

gg_resid

gg_resid + 
  geom_vline(xintercept = heights_aug_res_summary$values,
             colour = "salmon") + 
  ggrepel::geom_label_repel(data = heights_aug_res_summary,
             mapping = aes(label = names,
                           x = values,
                           y = y_pos))
```

The histogram of residuals from the model is seen in figure XX. The overlayed five number summary shows the minimum, 1st quantile, median, 3rd quantile, and maximum residual value residuals. To understand these residuals it can be useful to find those observations in the model that best match to this five number summary. This can be achieved using `keys_near` in `brolgar`. By default this links back to the 5 number summary, but any function can be used. You specify the variable you want to find the keys nearest, in this case `res`, the residuals:

```{r heights-aug-keys-near}
keys_near(heights_aug, 
          var = res)
```

To plot the data, they need to be joined back to the original data, and can then be plotted like so:

```{r}
library(stickylabeller)

heights_near_aug <- keys_near(heights_aug, 
          var = res) %>% 
  left_join(heights_aug,
            by = c("country"))

ggplot(heights_near_aug,
       aes(x = year,
             y = pred,
             group = country,
             colour = country)) + 
  geom_line() + 
  geom_point(aes(y = height_cm)) + 
  facet_wrap(~stat + country,
             labeller = label_glue("Country: {country} \nNearest to {stat} residual")) + 
  theme(legend.position = "none")
```

From observing this, we see that the minimum residual for Ireland has a few outliers that throw the result off, and that perhaps a different model with a non-linear term would be useful for improving the model fit.

More fits can be added, by changing `top_n = 5`, allowing for more 

```{r}
heights_near_aug <- keys_near(heights_aug, 
          var = res,
          top_n = 5) %>% 
  left_join(heights_aug,
            by = c("country"))

ggplot(heights_near_aug,
       aes(x = year,
             y = pred,
             group = country,
             colour = country)) + 
  geom_line() + 
  geom_point(aes(y = height_cm)) + 
  facet_wrap(~stat + country,
             labeller = label_glue("Country: {country} \nNearest to {stat} residual"),
             ncol = 5) + 
  theme(legend.position = "none")
```


# How to install and get started

(note: take a look at existing R package journal papers to see these examples)
(note: overview of function structure)
(goal for next week - content finished and paring down of paper to reasonable length.)

# Summary 

Problem #1: How do I look at some of the data?
Longitudinal data as a time series 
Specify structure, get a free lunch.
Look at as much of the raw data as possible
Use facet_sample() / facet_strata()

Problem #2: How do I find interesting observations?
Decide what features are interesting
Summarise down to one observation per key
Decide how to filter
Join this feature back to the data

Problem #3: How do I understand a model?
Look at (one, more or all!) subsamples
Arrange subsamples
(actually use similar approaches to earlier!)

Future work

* more features and stratification processes
* generalise to use for other structured data that doesn't have time components

# References {-}

<div id="refs"></div>

# Acknowledgements {-}

Tania Prvan
Stuart Lee
Mitchell O'Hara Wild
Earo Wang
Rob Hyndman
Miles McBain
Monash University
ACEMS