# Planning stuff

  * Issues with longitudinal data
    * spaghetti plots + hierarchical modelling
    * we lose information on the individual
  * identifying the individual experience amongst the collective mass is hard FIGURE - SPAGHETTI PLOT
    * why is it hard?
    * we don't see the individual
    * overplotting
    * identifying interesting observations or useful subplot sizes increases chances of a **distraction threshold** being reached
      * transparency doesn't really help
    * identifying a subplot size
    * Because identifying what is "interesting" involves a few steps that can easily be lost as the problem is dealt with.
* We need methods to understand the individuals and the individual experience
* `brolgar` provides a set of methods for exploring individual experience in 
  longitudinal data.
* The paper proceeds in four sections
  1. Mechanics
  1. How to look at sample of the data
  1. How to identify interesting observations
  1. how to understand a model
1. Mechanics
    1. Data structures (`tsibble`) - from ["data structures" vignette](http://brolgar.njtierney.com/articles/longitudinal-data-structures.html) - @njtierney 
        - what is longitudinal data
    1. Calculations (`features` + `key_slope`) []
    1. [Calculating Features](http://brolgar.njtierney.com/articles/finding-features.html) - @njtierney 
       - outline the different things we can create.
       - explain what can be calculated, and what is available that is being calculated.
       - How to create your own (e.g., mean(diff(x) as opposed to `increasing`
       - (1. [Exploratory Modelling](http://brolgar.njtierney.com/articles/exploratory-modelling.html) - @njtierney (really this is identifying features - since slope is just another feature that we can use, reduce this section / find overlap with "finding features").)
1. Explorations
    1. [Finding interesting observations](http://brolgar.njtierney.com/articles/id-interesting-obs.html) - @njtierney 
    1. [Summarising individuals]() - keys_near and colour palettes (use of `palap`).
    1. [Diagnosing models](http://brolgar.njtierney.com/articles/mixed-effects-models.html) - @njtierney 
1. Summary + Discussion

# Summary

# Misc notes

- Provide some terminology notation in the data structures part, so we can refer to them easier in the paper
  - Terminology: `index`, `key`, `features`, `multivariate`, `data`
  - Notation - $x_{it}$ ,$w_i$
      - data = $x$ (data can be multivariate)
      - features = ($w$) - (features could also be multivariate)
      - $i$ = individuals, the `key` - (i = 1:n).
      - $t$ = time, the `index`, (t = 1:n_{i} - the number of observations per individual) 
  - Identifying the overall data structure
     - long form for the longitudinal data (multiple rows per key)
     - wide form for features - (one row per key, one column for each feature)

 with a negative slope using `gghighlight`:

```{r use-gg-highlight}
library(gghighlight)

wages_slope %>% 
  as_tibble() %>% # workaround for gghighlight + tsibble
  ggplot(aes(x = xp, 
             y = ln_wages, 
             group = id)) + 
  geom_line() +
  gghighlight(.slope_xp < 0)
```


Specifically, we address: How to look at samples of the data? How to find interesting samples? How to find interesting individuals, finding who has the greatest change in height? Which has the tallest or shortest people? Which are most similar to the minimum, maximum, and median?


# plotting stuff for features

These could then be used to identify individuals who only increase like so:

```{r heights-mono-filter}
library(dplyr)
heights_mono %>%
  filter(increase)
```

They could then be joined back to the data

```{r wages-mono-join}
heights_mono_join <- heights_mono %>%
  filter(increase) %>%
  left_join(heights, by = "country")

heights_mono_join
```

And these could be plotted:

```{r gg-wages-mono}
ggplot(heights_mono_join,
       aes(x = year,
           y = height_cm,
           group = country)) + 
  geom_line()
```

To get a sense of the data and where it came from, we could create a plot with `gghighlight` to highlight those that only increase, by using `gghighlight(increase)` - since `increase` is a logical, this tells `gghighlight` to highlight those that are TRUE. 

```{r gg-high-mono}
library(gghighlight)
wages_mono %>%
  left_join(wages, by = "id") %>%
  ggplot(aes(x = xp,
             y = ln_wages,
             group = id)) +
  geom_line() + 
  gghighlight(increase)
```

You can explore the available features, see the function [References](http://brolgar.njtierney.com/reference/index.html)
